\large{\bf Quantitative genetics with stable laws }

### 2023-06-06 writeup.tex
one more section
snp effects code

\documentclass{article}
\usepackage{fullpage}
\usepackage{amsmath, amssymb}
\usepackage[hidelinks]{hyperref}
\usepackage{natbib}

@@

@@

It is mathematically intriguing to speculate what effect a non-Gaussian distribution
for $\epsilon$ would have on the dynamics induced by~\eqref{eqn:offspring_model}.
But, is there any evidence that non-Gaussian segregation noise is important in practice?
In this paper we focus primarily on the theory,
but first take a brief and imperfect look at what some data have to say on the topic.
To do this, we downloaded the GWAS results provided by~\citet{biobankSNPs}
for 7,221 phenotypes from 361,194 individuals from the March 2018 release of the UK Biobank.
Taken at face value, these provide for each phenotype a set of estimates of the additive effect
of the alternative allele at a set of SNPs on the phenotype relative to the reference allele.
The phenotypes we consider were all coded as 0/1,
so estimates could be thought of as additive adjustments to the risk
(so, \emph{not} on a logit scale).
These estimates were obtained as the coefficient for the SNP in a simple linear model
fit to each phenotype for each SNP,
with the first 20 principal components and all combinations of age, age squared, and inferred sex as covariates.
There are a great many possible issues with these data;
however, the results seem consistent across many SNPs and phenotypes,
and we know of no reason that possible methodological artifacts
(e.g., confounding by uncorrected variables)
would specifically affect the decay of the tail of effect sizes.

Suppose that the phenotype of each individual is additive: $X_f = \sum_i a_i$ and $X_m = \sum_i b_i$, for instance,
where each $a_i = 0$ if the individual is homozygous for the reference allele,
and that each contribution is inherited by the offspring with probability 1/2 (as if these were additive effects of rare, unlinked SNPs).
Then we would have the deviation of the offspring phenotype
from the midparent value equal to
$$
\epsilon = \sum_i (\theta_i a_i + \eta_i b_i)/2,
$$
where each $\theta_i$ and $\eta_j$ are independent Bernoulli(1/2).
So, the appropriate central limit theorem for $\epsilon$
is determined by the distribution of $\theta a_i$,
i.e.,

%%%%%%%%%%%%%%%%%%%%%%
\section{Extending the infinitesimal model}

%%%%%%%%%%%%%%%%%%%%%%
\section{Stationary distributions}

%%%%%%%%%%%%%%%%%%%%%%
\section{Simulations}

>>>>

\documentclass{article}
\usepackage{fullpage}
\usepackage{amsmath, amssymb}
\usepackage[hidelinks]{hyperref}
\usepackage{natbib}
\usepackage{xcolor}

@@

\comment{TODO expand, put somewhere:}
There are several justifications for why a ``stable infinitesimal model'' might be useful:
(1) as a phenomenological model;
(2) if it matches empirical observations; or
(3) if there is a central limit theorem suggesting it would be a good approximation in a wide range of situations.

@@

It is mathematically intriguing to speculate what effect a non-Gaussian distribution
for $\epsilon$ would have on the dynamics induced by~\eqref{eqn:offspring_model}.
But, is there any evidence that non-Gaussian segregation noise is important in practice?
In this paper we focus primarily on the theory,
but first take a brief and imperfect look at what some data have to say on the topic.
To do this, we downloaded the GWAS results provided by~\citet{biobankSNPs}
for 7,221 phenotypes from 361,194 individuals from the March 2018 release of the UK Biobank.
Taken at face value, these provide for each phenotype a set of estimates of the additive effect
of the alternative allele at a set of SNPs on the phenotype relative to the reference allele.
The phenotypes we consider were all coded as 0/1,
so estimates could be thought of as additive adjustments to the risk
(so, effects are \emph{not} on a logit scale).
These estimates were obtained as the coefficient for the SNP in a simple linear model
fit to each phenotype for each SNP,
with the first 20 principal components and all combinations of age, age squared, and inferred sex as covariates.
There are a great many possible issues with these data;
however, the results seem consistent across many SNPs and phenotypes,
and we know of no reason that possible methodological artifacts
(e.g., confounding by uncorrected variables)
would specifically affect the decay of the tail of effect sizes.

Suppose that the phenotype of each individual is additive:
the maternal trait value is $x = \sum_i x_{i,1} + x_{i,2}$
and the paternal trait value is $y = \sum_i y_{i,1} + y_{i,2}$,
where $x_{i,1}$ and $x_{i,2}$ are the effects of the alleles carried
by the mother on each of the two chromosomes
at the $i^\text{th}$ locus, respectively.
Then the offspring's trait is $z = \sum_i z_{i,m} + z_{i,p}$,
where $z_{i,m} = \theta_{i,m} x_{i,1} + (1-\theta_{i,m}) x_{i,2}$,
and $\theta_{i,m} = 1$ with probability 1/2 and $\theta_{i,m} = 0$ otherwise;
and likewise $z_{i,p} = \theta_{i,p} y_{i,1} + (1-\theta_{i,p}) y_{i,2}$.
With this notation, the segregation noise $z - \frac{x+y}{2}$
is a sum of terms of the form
$$
    \left( \theta_{i,m} - 1/2 \right) x_{i,1}
    + \left( 1/2 - \theta_{i,m} \right) x_{i,2} 
$$
(and similar terms with $y$ in place of $x$).
Now let $X$ be the effect of a uniformly chosen allele carried by a uniformly chosen individual
at a uniformly chosen locus.
The appropriate central limit theorem is determined by the tail behavior of $|(\theta - 1/2) X| = |X|/2$;
i.e., we'd like to know
$$
    \lim_{t \to \infty} \frac{ \log \P\{ |X| > 2 t \} }{ \log t } = \alpha ,
$$
if the limit exists.
Let $p_i$ be the minor allele frequency at the $i^\text{th}$ locus;
since the data we have assigns effect 0 to the major allele,
$x_i = 0$ with probability $1-p_i$.
Therefore, if we have estimated effects for $N$ SNPs
with minor allele frequency $p_i$ and effect size $e_i$ for the minor allele,
for $1 \le i \le N$,
then
$$
    \P\{ |X| > 2t \} \approx \frac{1}{N} \sum_{i : |e_i| > 2t} p_i .
$$
So, to obtain a per-trait estimate of $\alpha$,
we plot $\frac{1}{N} \sum_{i : |e_i| > t} p_i$ against $t$
for a sequence of values of $t$,
and fit a linear model to the tail
(which we take to be all values of $t$ for which the number of SNPs with absolute effect above $t$
is between 20 and 10\% of all SNPs).
Examples are shown in figure \comment{XXX}.

%%%%%%%%%%%%%%%%%%%%%%
\section{Extending the infinitesimal model}  % TODD

%%%%%%%%%%%%%%%%%%%%%%
\section{Convergence to the non-infinitesimal model}  % TODD

%%%%%%%%%%%%%%%%%%%%%%
\section{Stationary distributions}  % TODD

%%%%%%%%%%%%%%%%%%%%%%
\section{Simulations}  % PETER
